HiC Contact Map Comaprison Using Graphlet Approach

### 2018-07-11 papers/manuscripts/manuscript.tex
mardas

\begin{abstract}

In this study, we investigated
dissimilarities between normal cells and leukemic cells
in terms of their three dimensional conformation.
We first thresholed HiC data corresponding to
one normal cell line and four leukemic cells lines.
We then used the thresholded data to extract the 
first 72 orbits for all loci in each contact map.
We then measured pairwise graphlet distances as
well as pairwise graphlet distribution correlations
for each pair of cells and compared them using 
statistical methods.
Our results show that normal-cancer pairs have
significantly higher dinstances from each other
than cancer-cancer pairs.
Also, we concluded that for certain orbits,
cancer-cancer pairs demonstrate higher correlations
than normal-cancer pairs in terms of orbit frequency
distributions.

\end{abstract}

\section{Introduction}
This reasearch is an effort to compare the three-dimensional
structure of cell genomes. To the best of our knowledge
little research has dealt with quantitative comparison
of spatial conformation of chromosomes.
Ideally, it is desirable to compare 3D structures of 
cell in order to make such comparisons.
However, the main challenge in this regard
is that 
3D structure of a cell is not readily available. 
In order to find dissimilarities in the 3D structure of 
chromosomes, we used HiC dataset.
The HiC method, which was developed by 
\cite{lieberman2009comprehensive}, captures interactions between 
chromosomal fragments in kilobase resolution. 
Based on HiC data, a \textit{contact map} or
\textit{interaction frequency (IF) } matrix can be developed 
between \textit{loci} at a desired resolution.
A cell $IF_{ij}$ in an interaction frequency matrix captures 
the number of interaction detected
in HiC dataset between locus $i$ and locus $j$ in the genome.
An interaction matrix can be used to develop both 
inter- and intra-chromosomal interaction matrices.
We have found statistically significant
differences between interaction matrices of
normal cells and cancerous ones.

Recently, HiC data has been mostyly used
in efforts to predict the 3D conformation
of chromosomes using HiC data. 
\cite{noble2011three, 
rousseau2011three, 
hu2013bayesian, 
varoquaux2014statistical, 
trieu2014large, 
zhang20133d, 
lesne20143d, 
bau2011three, 
adhikari2016chromosome3d}.
These efforts, although different in
approach, usually translate 
interaction frequencies in contact maps as an inverse measure of 
distance. For example,  \cite{adhikari2016chromosome3d} developed
a method, Chromosome3D which uses
distance geometry simulated annealing (DGSA) in order to
reconstruct genome structure. These reconstructed 3D 
structures could be used in order to compare normal and malignant
cell lines. However, the main obstacle to doing so is that the
process of reconstruction is computationally
expensive. Almost all reconstruction methods rely on either 
deterministic or heuristic optimization approaches
in order to find the best
confomation that matches contact maps. Also, there is no real 
way of verifying whether the reconstruction results are accurate.
The only method of verifying such results is
Fluorescence in situ hybridizaiton
(FISH) which can capture the actual
3D configuration of chromosomes.
\cite{gasser2002visualizing, lanctot2007dynamic, mateos2009spatially}
However, this method can only be used locally and cannot map
the whole structure of the chromosomes.

Recently, graphlet comparison has emerged
as a novel method for comparing
large networks in order to
find local similarities in them.
Authors of \cite{prvzulj2007biological} 
provide a new measure of protein-protein
interaction (PPI)
network comparison
based on 73 constraints. The method
is used in order to compare two large
PPI networks in order to detect similarities.

In \cite{milenkoviae2008uncovering} the authors
 provide heuristics to compare two nodes in a graph
based on signature vectors, which are 73-dimensional vectors
$\mathbf{s}^T
= [s_0, s_2, ..., s_{72}]$ where $s_i$ denotes the number of nodes in
the network that are part of an orbit $i$. \\
They concluded that proteins with similar surroundings perform
similar functions.

In \cite{milenkovic2010cancer}, the same author investigates 
cancer-causing genes to find similarities in their signatures. After
clustering the genes based on \textit{signature similarity} criteria,
some clusters contain a lot of cancerous genes.
They use 4 different clustering methods with varying parameters to cluster
the proteins. They then predict the cancer-relatedness of a protein 
$i$ using
an enrichment criteria $\frac{k}{|C_i|}$ where $C_i$ is the cluster
where protein $i$ belongs and $k$ is the number of cancer-causing
proteins in $C_i$ and $|C_i|$ is the size of $C_i$.

In this paper, we applied the same approach of graphlet distance
and graphlet distribution comparison on HiC data
in order investigate genome spatial confomations differences
between normal and leukemic cells.

% **************************************************************
% Keep this command to avoid text of first page running into the
% first page footnotes
\enlargethispage{-65.1pt}
% **************************************************************

\subsubsection{Notations}
In this paper, matrices and vectors are represented using bold
capital and bold small letters respectively.
Matrix rows and columns are represented by a \textit{dot}
notation. For example, the $i$th row of matrix $M$ is
denoted by $M_{i.}$ and its $j$th column is represented
by $M_{.j}$.

@@

\section{Materials and Methods}
We re-used Leukemic Hi-C libraries 
created in \cite{wang2013properties}
These libraries were sequenced 
for cases of primary human B-acute
lymphoblastic leukemia (B-ALL or ALL), 
the MHH-CALL-4 B-ALL cell
line (CALL4), 
and the follicular lymphoma cell-line (RL).
As in \cite{wang2013properties}, 
We used normal B-cell line (GM068990)
from \cite{lieberman2009comprehensive} 
for our comparisons.
We created contact maps of resolution
500kb and normalized it using 
the \texttt{iced} package in python
developed by \cite{servant2015hic}.
We also performed and extra normalization by
performing network deconvolution \cite{feizi2013network}
in order to remove
the effect of indirect interactions.

>>>>

\begin{abstract}
    Chromosal Conformational Capturing (3C) data,
    particularly the Hi-C data is widely used for
    investigating the three-dimensional (3D) 
    chromatin conformation inside the nucleus and
    for identifying chromatin interactions as 
    well as topologically associating domains 
    (TADs). However, Hi-C based structural analysis
    for exploring local chromatin conformational 
    preferences has not been performed yet. 
    Here, for the first time, we present graphlet
    based local structural analysis for normal 
    and malignant human cell types  using chromosal
    conformational capturing (Hi-C) data. 
    We first applied thresholding in Hi-C interaction
    frequency data for one normal cell line and 
    four leukemic cells lines in order to covert 
    them into unweighted graphs. We then mined 
    the unweighted graphs to extract graphlet -based 
    orbitial frequency distributions? . Subsequently
    , pairwise graphlet distances as well as 
    pairwise graphlet distribution correlations 
    for each pair of cells were computed and 
    compared using statistical methods namely 
    ANOVA and multi-variate ANOVA. Our results 
    demonstrate that normal-cancer pairs have 
    significantly ($F(1, 1654) = 20.49$, $p-value
    < 0.0001$) higher difference than cancer-cancer 
    pairs. Furthermore, for certain orbits cancer-cancer 
    pairs demonstrated higher correlations 
    ($F(365, 7882)=391$, $p-value<0.00001$, and 
    $Wilk's \quad \Lambda=0456$) than normal-cancer pairs 
    in terms of orbit frequency distributions. 
    Source code of GrapHiClet 
    is freely available at 
    \url{https://github.com/rasoolianbehnam/watson}.
    Also, the database used for this project can be
    found at \url{https://graphiclet.localtunnel.me/secret/}
\end{abstract}

\section{Introduction}
Genome of a eukaryotic cell is organized into a
complex three-dimensional (3D) structure, which
leads to a network of chromosomal interactions \cite{dekker2008gene}.
The spatial organization of genome is one of
the most important factors in determining its
function \cite{fraser2007nuclear} via interactions between
spatially close genes \cite{kagey2010mediator}. Such
interactions can be modeled using a network approach
\cite{kagey2010mediator}. However, our knowledge of
3D conformation of genome is limited. Fluorescent In
Situ Hybridization (FISH) technology, which can
capture 3D conformation of multiple loci but cannot
capture global layout of the genome. In order to
address this, more advanced techniques of chromosome
conformation capturing such as 3C \cite{dekker2002capturing},
4C \cite{simonis2006nuclear} and 5C \cite{dostie2007mapping}
has been introduced which are based on chromatin fragment
fixation, using restriction-enzyme digestion and
intra-molecular ligation. One of the most recent
versions of chromosome conformation capturing is
Hi-C \cite{rao20143d, lieberman2009comprehensive},
which $IF_{ij}$ captures interactions between chromosomal
fragments resulting in a contact map or interaction frequency
(IF) matrix, which captures the number of interactions
detected in Hi-C dataset between each pair of loci.

Recently, Hi-C data has been mostly used to predict
the 3D structure at the chromosome level by satisfying
intra-chromosomal interactions or even at the
full genome by leveraging both inter- and intra-chromosomal
interactions
\cite{noble2011three, 
rousseau2011three, 
hu2013bayesian, 
varoquaux2014statistical, 
trieu2014large, 
zhang20133d, 
lesne20143d, 
bau2011three, 
adhikari2016chromosome3d}.
These efforts, although different in
approach, usually translate interaction frequencies in
contact maps as simple inverse measures of distance.
They then try to find the coordinates of the chromosomal
loci that would result in that particular distance
matrix using optimization. These methods vary in
the optimization models and methodologies used.
Also, efforts has been made recently to learn the
inverse conversion parameters from contact maps
to distance matrices \cite{oluwadare2018maximum}.
The reconstructed 3D structures could be used in
order to compare normal and malignant cell lines.
However, the main obstacle to doing so is that
the process of reconstruction is computationally expensive.
Almost all reconstruction methods rely on either
deterministic or heuristic optimization approaches in
order to find the best conformation that matches contact
maps. Also, there is no experimental way of evaluating
the accuracy of the reconstructed 3D conformations.
One experimental way of partially validating correctness
of reconstruction process is Fluorescence in
situ hybridizaiton (FISH) method. However, this
technique is limited to a few loci and cannot be
extended to cover the whole structure of the chromosomes.
Furthermore, the process of reconstructing 3D
chromatic structure is computationally expensive,
hindering their applicability at the genome-wide scale.
% **************************************************************
% Keep this command to avoid text of first page running into the
% first page footnotes
\enlargethispage{-65.1pt}
% **************************************************************

Over the past decade, graphs have been used extensively
in order to model domain-specific phenomena such
as computer security, gene regulatory networks and
social networks. These graphs are usually in
the form of very large networks. It is often necessary
to perform statistical analysis between graphs in
order to compare their structure. Graphlets and
orbits can be used in order to probe large graphs
in order to find global and local similarities 
\cite{shervashidze2009efficient, borgs2006counting,
bondy1977graph}.
This can be done by counting the number of occurrences
of a each graphlet and/or orbits for each node
in the whole graph and the comparing them \cite{prvzulj2007biological,
prvzulj2004modeling}

Here, we present local structural analysis between
4 sets of Hi-C data. All four data sets are sequences
from the same cell line, with one of them being
a normal cell and the other three sequenced from
three types of leukemic cells. In order to achieve
this, we used graphlets, given their strength in
capturing information about local structures of
a graph. We first thresholded Hi-C interaction frequency
matrices in order to convert in to an unweighted undirected
graph adjacency matrix. We the extracted counts of
the first 73 orbits for each node, identifying each
node with a signature vector of size 73. We then
applied graphlet distance metrics proposed in
{prvzulj2007biological} together with statistical methods
in order to find difference between cell lines.
Our results show that difference of local structure
between normal cells and leukemic cells are significantly
larger that difference between cancer-cancer cells.

\subsubsection{Notations}
%In this paper, matrices and vectors are represented using bold
%capital and bold small letters respectively.
%Matrix rows and columns are represented by a \textit{dot}
%notation. For example, the $i$th row of matrix $M$ is
%denoted by $M_{i.}$ and its $j$th column is represented
%by $M_{.j}$.

@@

\section{Materials and Methods}
Chromosomes inside the nuclei are made up of
pairs of nucleotides called a base. As a result of
the Hi-C process, a database is generated which
provides interaction counts found by the method between
a number of such bases. The minimum number of
bases that can be captured is called the resolution of
that database. These counts are then binned so
that counts are aggregated for every equal-sized
length of the chromosome (e.g. 1 Mb pairs) leading
to an N * N interaction frequency matrix. Each
cell (i, j) in the matrix is the aggregate count
of all interactions found between loci i\textsuperscript{th} and
j\textsuperscript{th} length.
We re-used Leukemic Hi-C libraries created in
\cite{wang2013properties}. These libraries were sequenced
using Illumina HiSeq 2000 for cases of primary human
B-acute lymphoblastic leukemia (B-ALL or ALL),
the MHH-CALL-4 B-ALL cell line (CALL4), and the
follicular lymphoma cell-line (RL) for which high-quality
paired-end reados of 39M, 79M and 33M were obtained
respectively As in \cite{wang2013properties}, We used
normal B-cell line (GM068990) from as benchmark for
our comparisons. The three datasets generated in
\cite{wang2013properties} were valid since 98\% of
the contact generated in \cite{wang2013properties}
were identical to that of \cite{lieberman2009comprehensive}
and 83\% of contacts in \cite{lieberman2009comprehensive}
were also present in \cite{wang2013properties}. We
created contact maps with bin sizes of  500 kilo-base
and normalized them using normalization provided 
HiC-Pro( \texttt{iced} package in python) developed
by \cite{servant2015hic}.
Normalization is necessary since Hi-C data usually
contains different biases due to GC content,
mappability and effective fragment length \cite{yaffe2011probabilistic,
hu2012hicnorm}. The normalization proivdedi in
the iced package is based on the Sinkhorn-Knopp
algorithm which is a simple, parameter-free and
capability to correct unknown biases.  The edges
in a network usually include indirect dependencies because
correlations are transitive; that is, if there is
a strong realationship between nodes 1 and 2,
and also a strong realationship between nodes 2
and 3, it is highly likely that nodes relationship between
nodes 1 and 3 is exaggerated in the network \cite{feizi2013network}.
In order to remove the effect of indirect interactions,
We also performed and extra normalization by
performing network deconvolution \cite{feizi2013network},
which uses eigenvalue decomposition and infinite series
sums in order to reverse the bias posed by indirect
relationships.

\subsection{Graphlets and Orbits}
\begin{figure}[H]
    \centering
    \includegraphics[width=.5\textwidth]{figures/graphlets.png}
    \caption{}
    \label{fig_graphlets}
\end{figure}
Recently, graphlet comparison has emerged as
a novel method for comparing large networks in
order to find local similarities in them. A
graph G is a pair $(V,E)$, where $V$ is a set of
vertices and $E \subseteq V\times V$ is a set of
edges. A connected graph is one where there is
a path between every pair of vertices.
Given a graph $G(V, E)$ and $S \subseteq V$,
then $G'(S, E')$ is a graphlet if and only if
it is connected $E' = \{(u, v) | u, v \in V \quad \&
\quad (u, v) \in E \rightarrow (u, v) \in E'\}$. 
There a total of 30 graphlets
of size 2, 3, and 5.  Figure \ref{fig_graphlets}
demonstrates these graphlets. 
As can be seen there is only one graphlet
of size 2 which is equivalent to an edge; that
is, the number of graphlet 1 in a graph is the
same as the number of edges on the graph. There
are also a total of 2 graphlets of size 2, 6
of size 4 and 20 of size 5. The nodes of each graphlet
can be partitioned into sets of topographically equivalent
nodes called orbits. For  example, in  Figure 
\ref{fig_graphlets}, we can see that G3 can be
partitioned into 2 sets of nodes, the middle ones
(white) and the outer nodes (black). 
For the same set of graphlets,
there are 73 orbits that are also illustrated in
figure  Figure using nodes of difference shades.
